---
title: "Competing Risks IPCW: Guided Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Competing Risks IPCW: Guided Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette demonstrates IPCW methods for competing risks data. We simulate
a prostate cancer dataset where metastasis (event_1) and death (event_2) are
competing risks, and PSA quartile at baseline (`z1`) both predicts metastasis
risk and drives informative censoring.

## Setup

```{r packages, message = FALSE, warning = FALSE}
library(ipcw)
library(survival)
library(ggplot2)
library(purrr)
```

## Simulate data

`sim_data_cr()` generates competing risks data under a sub-distribution hazard
model. The covariate `z1` is a four-level factor (PSA quartile). We use
baseline-dependent censoring so that censoring is informative.

```{r simulate}
set.seed(9843)
dat <- sim_data_cr(
  n          = 500,
  censoring  = "baseline",
  beta1      = log(1.5),
  beta2      = log(2.25),
  beta3      = log(3.4),
  p          = 0.3
)
table(dat$delta)
```

## Explore the data

```{r explore, fig.width = 6, fig.height = 4}
# Cumulative incidence by PSA quartile
plot(survfit(Surv(t, delta) ~ z1, data = dat),
     col = 1:4, lty = rep(1:2, each = 4),
     xlab = "Time", ylab = "Probability")
legend("topleft", legend = paste("z1 =", 0:3), col = 1:4, lty = 1, bty = "n")
```

```{r censoring-dist, fig.width = 6, fig.height = 4}
# Censoring distribution by PSA quartile
plot(survfit(Surv(t, delta == "censor") ~ z1, data = dat),
     col = 1:4, xlab = "Time", ylab = "P(not censored)")
legend("topright", legend = paste("z1 =", 0:3), col = 1:4, lty = 1, bty = "n")
```

## Convert to long format and add IPCW weights

```{r long-format}
dat_long <- wide_to_long_cr(dat)

# Cox model weights
dat_long_cox <- add_ipcw_weights_cr(dat_long, strat = "no")

# Stratified (non-parametric) weights
dat_long_strat <- add_ipcw_weights_cr(dat_long, strat = "yes")
```

```{r weight-histogram, fig.width = 5, fig.height = 3}
ggplot(dat_long_strat, aes(x = 1 / p_notcens)) +
  geom_histogram(bins = 40) +
  xlab("IPCW weight") + ylab("Count") +
  theme_bw()
```

## Estimate cumulative incidence

```{r cuminc}
esttimes <- sort(dat$t)

ci_naive   <- cuminc_naive_cr(dat, esttimes)
ci_wavg    <- cuminc_waverage_cr(dat, esttimes)
ci_cox     <- cuminc_ipcw_cr(dat_long_cox, esttimes)
```

```{r cuminc-plot, fig.width = 6, fig.height = 4}
to_plot <- rbind(
  data.frame(time = esttimes, est = ci_wavg, method = "Weighted average"),
  data.frame(time = esttimes, est = ci_cox,  method = "Cox IPCW")
)

ggplot(to_plot, aes(x = time, y = est, colour = method)) +
  geom_step() +
  xlim(0, 5) + ylim(0, 0.65) +
  xlab("Time") + ylab("Cumulative incidence of event 1") +
  theme_bw(base_size = 14) +
  theme(legend.position = "bottom", legend.title = element_blank())
```

## Fine-Gray IPCW regression

```{r fine-gray}
dat_long_fg <- fg_split_cr(dat_long)

# Stratified weights
dat_long_fg_strat <- add_fg_weights_cr(dat_long_fg, strat = "yes")
fg_strat <- fg_weighted_cr(dat_long_fg_strat, extend = FALSE)
exp(fg_strat[, 1])   # exponentiated coefficients

# Cox model weights
dat_long_fg_cox <- add_fg_weights_cr(dat_long_fg, strat = "no")
fg_cox <- fg_weighted_cr(dat_long_fg_cox)
exp(fg_cox[, 1])
```

## Bootstrap standard errors

Bootstrap resampling provides valid standard errors when the weights are
estimated. The example below uses 500 samples; run in parallel as needed.

```{r bootstrap, eval = FALSE}
set.seed(20240917)

boot_dat      <- map(1:500, ~ slice_sample(dat, prop = 1, replace = TRUE))
boot_dat      <- map(boot_dat, function(x) { x$id <- seq_len(nrow(x)); x })
boot_dat_long <- map(boot_dat, wide_to_long_cr)

# Weighted-average cumulative incidence bootstrap
w_avg_boot <- map(boot_dat, ~ cuminc_waverage_cr(., esttimes))
w_avg_boot <- matrix(unlist(w_avg_boot), ncol = length(esttimes), byrow = TRUE)
lower_wavg  <- apply(w_avg_boot, 2, function(x)
                 ifelse(any(is.na(x)), NA, quantile(x, 0.025)))
upper_wavg  <- apply(w_avg_boot, 2, function(x)
                 ifelse(any(is.na(x)), NA, quantile(x, 0.975)))

# Cox IPCW cumulative incidence bootstrap
cox_boot      <- map(boot_dat_long, ~ cuminc_ipcw_cr(add_ipcw_weights_cr(., strat = "no"), esttimes))
cox_boot      <- matrix(unlist(cox_boot), ncol = length(esttimes), byrow = TRUE)
lower_cox     <- apply(cox_boot, 2, function(x)
                   ifelse(any(is.na(x)), NA, quantile(x, 0.025)))
upper_cox     <- apply(cox_boot, 2, function(x)
                   ifelse(any(is.na(x)), NA, quantile(x, 0.975)))

# Fine-Gray stratified bootstrap
boot_dat_fg   <- map(boot_dat_long, fg_split_cr)
fg_strat_boot <- map(boot_dat_fg, ~ fg_weighted_cr(add_fg_weights_cr(., strat = "yes"),
                                                 extend = FALSE)[, 1])
fg_strat_boot <- matrix(unlist(fg_strat_boot), ncol = 3, byrow = TRUE)
lower_fg_strat <- apply(fg_strat_boot, 2, function(x)
                     ifelse(any(is.na(x)), NA, quantile(x, 0.025)))
upper_fg_strat <- apply(fg_strat_boot, 2, function(x)
                     ifelse(any(is.na(x)), NA, quantile(x, 0.975)))

exp(fg_strat[, 1])
exp(lower_fg_strat)
exp(upper_fg_strat)
```
