---
title: "Single-Event IPCW: Guided Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Event IPCW: Guided Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
has_ggsurvfit <- requireNamespace("ggsurvfit", quietly = TRUE)
has_gt        <- requireNamespace("gt",        quietly = TRUE)
```

This vignette demonstrates inverse probability of censoring weighting (IPCW)
for a single-event survival outcome. We use a simulated dataset that mimics a
clinical trial of patients with chronic myeloid leukemia. The primary endpoint 
is progression-free survival (PFS). Participants are randomized to receive 
either chemotherapy or a tyrosine kinase inhibitor (TKI). Participants can be 
censored for the primary endpoint due to treatment intolerance or excessive
toxicity, which is expected to occur equally for the two treatment groups.
Treatment group is the factor of interest in this study. Dependent
censoring arises because participants with higher absoluate neutrophil count 
(ANC) at baseline are less likely to go off study due to treatment intolerance,
and therefore to be censored for PFS, and are also less likely to experience a
progression or die.

## Setup

```{r packages, message = FALSE, warning = FALSE}
library(ipcw)
library(survival)
library(purrr)
```

## Simulate data and compute IPCW weights

First, create a simulated dataset of `n=500` patients using the `sim_data_se()` function. Key parameters include:

- `t`: observed time (event or censoring)
- `delta`: event indicator (1 = event, 0 = censored)
- `x`: treatment (0 = chemotherapy, 1 = TKI)
- `W2`: ancillary biomarker that drives informative censoring (ANC)

```{r generate_data}
set.seed(20240429)
dat <- sim_data_se(n = 500)
```

`get_ipcw_wgt()` converts the wide dataset to counting-process (long) format
and appends the unstabilized IPCW weight column `wgt`. 

```{r compute-weights-live}
dat_long <- get_ipcw_wgt(dat)
```

## IPCW Kaplan-Meier curves

The chunk below requires the **ggsurvfit** package. Install it from CRAN if it 
is not already installed. We plot the IPCW Kaplan-Meier estimates.

```{r km-plot, fig.width = 6, fig.height = 4}
# install.packages("ggsurvfit")
library(ggsurvfit)

km_fit <- 
  survfit2(
    Surv(tstart, tstop, delta) ~ x,
    data = dat_long,
    weights = wgt,
    timefix = FALSE)

ggsurvfit(km_fit) +
  xlim(c(0, 1000)) +
  scale_color_hue(labels = c("Chemotherapy", "TKI")) +
  labs(
    x = "Days from treatment start",
    y = "Progression-free survival probability"
  )
```

```{r}
B <- 500

set.seed(20240917)

ipcw_boot_dat <- 
  map(
    1:B,
    ~ slice_sample(dat, prop = 1, replace = T)
  ) |> 
  map(~ get_ipcw_wgt(.))

# Fit the IPCW K-M curves to the bootstrapped data
ipcw_km_boot_fit <- 
  ipcw_boot_dat |> 
  map(~get_ipcw_km_prob_x(., pre_times = seq(0, 2429, 1)))

# Collapse into a single dataframe
ipcw_km_boot_fit_all <- 
  ipcw_km_boot_fit |>  
  map_dfr(bind_rows, .id = "sim_num")

# Then summarize at each time for each treatment
# And get the 2.5 and 97.5 percentiles
ipcw_km_ci_summ <-
  ipcw_km_boot_fit_all |> 
  group_by(time, x) |> 
  summarize(
    lpci_surv = quantile(surv, probs = 0.025),
    upci_surv = quantile(surv, probs = 0.975),
    median_surv = median(surv),
    mean_surv = mean(surv)
  )

# Add in the original point estimates based on the single dataset
point_orig <- summary(
  survfit2(Surv(tstart, tstop, delta) ~ x, 
           data = dat_long, weights = wgt, timefix = FALSE) ,
  times = seq(0, 2429, 1)
  )

point_orig_df <- 
  tibble(
    x = point_orig$strata,
    time = point_orig$time,
    surv = point_orig$surv
  ) |> 
  mutate(
    x = as.factor(x)
  )

ipcw_km_ci_summ_orig <- 
  ipcw_km_ci_summ |> 
  full_join(point_orig_df)

ggplot(ipcw_km_ci_summ_orig, aes(x = time, color = x, fill = x)) +
  geom_step(aes(y = surv)) +
  geom_ribbon(aes(ymin = lpci_surv, ymax = upci_surv), 
              alpha = 0.3, linetype = "dashed") +
  labs(x = "Time", y = "Survival Probability") +
  theme_minimal() + 
  xlim(c(0, 1000)) +
  labs(
    x = "Days from treatment start",
    y = "Progression-free survival probability"
  ) +
  theme(legend.position = "bottom",
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 6),
        legend.text = element_text(size = 6))
```



## IPCW Cox regression

```{r cox-fit}
ipcw_cox_fit <- coxph(
  Surv(tstart, tstop, delta) ~ x + cluster(id),
  data    = single_example_ipcw_dat,
  weights = single_example_ipcw_dat$wgt,
  timefix = FALSE
)
summary(ipcw_cox_fit)$coef
```

Or use the convenience wrapper, which also returns the hazard ratio and
robust-SE-based confidence interval:

```{r cox-fit-wrapper}
get_ipcw_cox_fit(single_example_ipcw_dat, weight = "wgt")
```

## Bootstrap percentile confidence intervals

Bootstrap resampling is needed for valid confidence intervals because the
weights are estimated. The code below runs 500 bootstrap samples; in practice
you may want to run this in parallel.

```{r bootstrap, eval = FALSE}
set.seed(20240917)
B <- 500

ipcw_boot_dat <- map(1:B, ~ dplyr::slice_sample(single_example_dat,
                                                  prop = 1, replace = TRUE))
ipcw_boot_dat <- map(ipcw_boot_dat, get_ipcw_wgt)

ipcw_cox_boot <- map(ipcw_boot_dat, ~ get_ipcw_cox_fit(., "wgt"))
boot_log_hrs  <- map_dbl(ipcw_cox_boot, "log_hr")

boot_pci <- get_boot_pci(data.frame(log_hr = boot_log_hrs))

result_tab <- data.frame(
  Scale          = c("log(HR)", "HR"),
  Estimate       = c(coef(ipcw_cox_fit), exp(coef(ipcw_cox_fit))),
  Lower_95pct_PI = c(boot_pci[[1]], exp(boot_pci[[1]])),
  Upper_95pct_PI = c(boot_pci[[2]], exp(boot_pci[[2]]))
)
print(round(result_tab[, -1], 2))
```
