---
title: "Single-Event IPCW: Guided Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Event IPCW: Guided Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
has_ggsurvfit <- requireNamespace("ggsurvfit", quietly = TRUE)
has_gt        <- requireNamespace("gt",        quietly = TRUE)
```

This vignette demonstrates inverse probability of censoring weighting (IPCW)
for a single-event survival outcome. We use a simulated dataset that mimics a
clinical trial comparing two treatments (chemotherapy vs. TKI) on
progression-free survival, where censoring is informative because it depends
on an ancillary biomarker (`W2`).

## Setup

```{r packages, message = FALSE, warning = FALSE}
library(ipcw)
library(survival)
library(purrr)
```

## Data

The package ships with `single_example_dat`, a simulated dataset of 500
subjects generated with `set.seed(20240429)`. The key variables are:

- `t`: observed time (event or censoring)
- `delta`: event indicator (1 = event, 0 = censored)
- `x`: treatment (0 = chemotherapy, 1 = TKI)
- `W2`: ancillary biomarker that drives informative censoring

```{r load-data}
data(single_example_dat)
head(single_example_dat)
```

The code below re-creates this dataset from scratch, so you can see exactly
how it was generated.

```{r generate-data, eval = FALSE}
# Fix simulation parameters
n       <- 500
alpha   <- 0.05
x_prop  <- 0.5
a       <- 2
sigma   <- 500
beta    <- log(0.25)
lambda  <- 0.01
phi     <- -5

set.seed(20240429)

Y1 <- rexp(n, rate = 1)
Y2 <- rexp(n, rate = 1)
Z  <- rgamma(n, shape = alpha, rate = 1)
W1 <- (1 + Y1 / Z)^(-alpha)
W2 <- (1 + Y2 / Z)^(-alpha)
x  <- rbinom(n, 1, prob = x_prop)

S     <- sigma * (-log(1 - W1) / exp(beta * x))^(1 / a)
C     <- rexp(n, rate = lambda * exp(phi * W2))
t     <- pmin(S, C)
delta <- as.integer(S <= C)

single_example_dat <- tibble::tibble(S, t, delta, x, W2)
```

## Prepare data and compute IPCW weights

`get_ipcw_wgt()` converts the wide dataset to counting-process (long) format
and appends the unstabilized IPCW weight column `wgt`. The package also
provides `single_example_ipcw_dat` with these weights pre-computed.

```{r load-ipcw-data}
data(single_example_ipcw_dat)
head(single_example_ipcw_dat)
```

Alternatively, compute the weights directly:

```{r compute-weights-live, eval = FALSE}
dat_long <- get_ipcw_wgt(single_example_dat)
```

## IPCW Kaplan-Meier curves

The chunk below requires the **ggsurvfit** package. If it is not installed,
you can plot the IPCW K-M curves with base R instead (see the commented code).

```{r km-plot, eval = has_ggsurvfit, fig.width = 6, fig.height = 4}
library(ggsurvfit)
km_fit2 <- survfit2(Surv(tstart, tstop, delta) ~ x,
                    data    = single_example_ipcw_dat,
                    weights = single_example_ipcw_dat$wgt,
                    timefix = FALSE)
ggsurvfit(km_fit2) +
  xlim(c(0, 1000)) +
  scale_color_hue(labels = c("Chemotherapy", "TKI")) +
  labs(
    x = "Days from treatment start",
    y = "Progression-free survival probability"
  )
```

```{r km-plot-base, eval = !has_ggsurvfit, fig.width = 6, fig.height = 4}
km_fit <- survfit(Surv(tstart, tstop, delta) ~ x,
                  data    = single_example_ipcw_dat,
                  weights = wgt,
                  timefix = FALSE)
plot(km_fit, col = 1:2, xlim = c(0, 1000),
     xlab = "Days from treatment start",
     ylab = "Progression-free survival probability")
legend("topright", legend = c("Chemotherapy", "TKI"), col = 1:2, lty = 1)
```

## IPCW Cox regression

```{r cox-fit}
ipcw_cox_fit <- coxph(
  Surv(tstart, tstop, delta) ~ x + cluster(id),
  data    = single_example_ipcw_dat,
  weights = single_example_ipcw_dat$wgt,
  timefix = FALSE
)
summary(ipcw_cox_fit)$coef
```

Or use the convenience wrapper, which also returns the hazard ratio and
robust-SE-based confidence interval:

```{r cox-fit-wrapper}
get_ipcw_cox_fit(single_example_ipcw_dat, weight = "wgt")
```

## Bootstrap percentile confidence intervals

Bootstrap resampling is needed for valid confidence intervals because the
weights are estimated. The code below runs 500 bootstrap samples; in practice
you may want to run this in parallel.

```{r bootstrap, eval = FALSE}
set.seed(20240917)
B <- 500

ipcw_boot_dat <- map(1:B, ~ dplyr::slice_sample(single_example_dat,
                                                  prop = 1, replace = TRUE))
ipcw_boot_dat <- map(ipcw_boot_dat, get_ipcw_wgt)

ipcw_cox_boot <- map(ipcw_boot_dat, ~ get_ipcw_cox_fit(., "wgt"))
boot_log_hrs  <- map_dbl(ipcw_cox_boot, "log_hr")

boot_pci <- get_boot_pci(data.frame(log_hr = boot_log_hrs))

result_tab <- data.frame(
  Scale          = c("log(HR)", "HR"),
  Estimate       = c(coef(ipcw_cox_fit), exp(coef(ipcw_cox_fit))),
  Lower_95pct_PI = c(boot_pci[[1]], exp(boot_pci[[1]])),
  Upper_95pct_PI = c(boot_pci[[2]], exp(boot_pci[[2]]))
)
print(round(result_tab[, -1], 2))
```
